---
title: "Game Analytics: From Exploratory Data Analysis to Predictive Modeling"
author: "Hoang Son Lai"
date: "Nov 17, 2025"

format:
  html:
    output-file: game_analytics.html
    css: "assignment.css"
    embed-resources: true
    code-fold: true
    toc: true
    toc-depth: 4
    fig-width: 6
    fig-height: 4
    fig-align: center
    fig-retina: 4

execute:
  echo: true
  message: false
  warning: false
  cache: false
  freeze: false
---

```{r}
#| echo: false
# Load libraries
library(tidyverse)
library(conflicted)
library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(knitr)
library(scales)
library(naniar)
library(lubridate)
library(tidytext)
library(viridis)
library(plotly)
library(patchwork)
library(purrr)
library(gt)
library(DT)
library(kableExtra)
library(stringr)
library(tidyr)
library(forcats)
library(caret)
library(randomForest)
library(cluster)
library(factoextra)
library(corrplot)
library(ggpubr)
library(RColorBrewer)
library(broom)
library(colorspace)
library(GGally)
library(tibble)
library(highcharter)
library(htmltools)
library(jsonlite)
library(ggrepel)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)
conflicts_prefer(plotly::layout)
conflicts_prefer(ggplot2::margin)

# Set random seed for reproducibility
set.seed(123)
```

## 1. Data Overview & Processing

```{r}
# Load and clean the data
game_data <- read.csv("data/game_sessions.csv", stringsAsFactors = FALSE)

# Data cleaning and preprocessing
game_data_clean <- game_data %>%
  mutate(
    start_time = as.POSIXct(start_time, format = "%Y-%m-%dT%H:%M:%OSZ"),
    end_time = as.POSIXct(end_time, format = "%Y-%m-%dT%H:%M:%OSZ"),
    death_reason = as.factor(death_reason),
    # Handle missing end_time
    game_duration = ifelse(is.na(game_duration), 0, game_duration),
    # Create performance metrics
    score_per_second = ifelse(game_duration > 0, score / game_duration, 0),
    accuracy = ifelse(bullets_fired > 0, ufos_shot / bullets_fired, 0)
  ) %>%
  filter(!is.na(start_time))  
```

```{r}
# Display basic information
cat("Cleaned Dataset Dimensions:", dim(game_data_clean), "\n")
cat("Date Range:", as.character(min(game_data_clean$start_time)), "to", 
    as.character(max(game_data_clean$start_time)), "\n")
```

```{r}
head(game_data_clean)
```

```{r}
#| label: tbl-1-1
#| tbl-cap: ""

variable_description <- tibble(
  Variable = c(
    "id",
    "start_time",
    "end_time",
    "score",
    "coins_collected",
    "ufos_shot",
    "bullets_fired",
    "death_reason",
    "game_duration",
    "pipes_passed",
    "score_per_second",
    "accuracy"
  ),
  Description = c(
    "Unique session identifier",
    "Timestamp when the game session started",
    "Timestamp when the game session ended",
    "Final score achieved in the session. Score = coins_collected + (3 ร ufos_shot)",
    "Number of coins collected by the player",
    "Number of UFO enemies shot",
    "Total number of bullets fired",
    "Cause of death (collision type / hazard)",
    "Total session duration in seconds",
    "Number of pipes the player successfully passed",
    "Score normalized by session duration (score รท game_duration)",
    "Shooting accuracy (ufos_shot รท bullets_fired)"
  ),
  Type = c(
    "Character",
    "Datetime",
    "Datetime",
    "Integer",
    "Integer",
    "Integer",
    "Integer",
    "Categorical",
    "Numeric",
    "Integer",
    "Numeric",
    "Numeric"
  )
)

variable_description %>%
  gt() %>%
  tab_header(
    title = md("**Variable Description - Plane Game Analytics**")
  ) %>%
  cols_width(
    Variable ~ px(160),
    Description ~ px(420),
    Type ~ px(120)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )

```

```{r}
#| label: tbl-1-2
#| tbl-cap: "Number of NA values per column in game_data_clean"

# Check NA
na_table <- game_data_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Column", values_to = "NA Count") %>%
  arrange(desc(`NA Count`))

# Display by kable
kable(na_table)
```

```{r}
#| label: tbl-1-3
#| tbl-cap: "Summary statistics for key variables"

# Summary statistics
game_data_clean %>%
  select(score, game_duration, coins_collected, ufos_shot, bullets_fired, pipes_passed, accuracy) %>%
  summary() %>%
  kable()
```

## 2. Exploratory Data Analysis

### 2.1. Distribution Overview

```{r}
#| fig-width: 8
#| fig-height: 7
#| fig-cap: "Distribution of Game Performance Metrics"
#| label: fig-2-1-1

# Tidy dataset long & clean variable names
game_long <- game_data_clean %>%
  select(score, game_duration, accuracy, coins_collected, ufos_shot) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    variable = str_replace_all(variable, "_", " "),  
    variable = str_to_title(variable)                 
  )

# Facet color
facet_colors <- c(
  "Score"            = "#4e79a7",
  "Game Duration"    = "#f28e2b",
  "Accuracy"         = "#59a14f",
  "Coins Collected"  = "#e15759",
  "Ufos Shot"        = "#76b7b2"
)

# Plot
ggplot(game_long, aes(x = value, fill = variable)) +
  geom_histogram(
    bins = 30,
    alpha = 0.7,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_manual(values = facet_colors) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "Distribution of Game Performance Metrics",
    x = "Value",
    y = "Count",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    strip.text         = element_text(face = "bold", size = 9),
    plot.title         = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.border       = element_rect(
      color = "white",
      fill  = NA,
      linewidth = 1
    ),
    axis.title.x       = element_text(size = 9, margin = margin(t = 5)),
    axis.title.y       = element_text(size = 9, margin = margin(r = 5)),
    axis.text.x        = element_text(size = 6),
    axis.text.y        = element_text(size = 6)
  )
```

```{r}
#| fig-height: 5
#| fig-cap: "Common Causes Of Death"
#| label: fig-2-1-2 

# Bar chart death reason
death_reason_data <- game_data_clean %>%
  count(death_reason, sort = TRUE) %>%
  mutate(
    death_reason_clean = death_reason %>%
      str_replace_all("_", " ") %>% 
      str_to_title()
  )

highchart() %>%
  hc_chart(type = "bar") %>%
  hc_title(text = "Common Causes Of Death") %>%
  hc_xAxis(
    categories = death_reason_data$death_reason_clean,
    title = list(text = "Death Reason"),
    labels = list(style = list(fontSize = "11px"))
  ) %>%
  hc_yAxis(
    title = list(text = "Number of games"),
    labels = list(style = list(fontSize = "11px"))
  ) %>%
  hc_add_series(
    name = "Count",
    data = death_reason_data$n,
    color = "#1f77b4"
  ) %>%
  hc_tooltip(
    useHTML = TRUE,
    pointFormat = "<b>Count:</b> {point.y}"    
  ) %>%
  hc_plotOptions(
    series = list(
      borderWidth = 0,
      dataLabels = list(enabled = FALSE)
    )
  ) %>%
  hc_legend(enabled = FALSE) %>%              
  hc_exporting(enabled = TRUE)
```

```{r}
#| fig-height: 5
#| label: fig-2-1-3
#| fig-cap: "Correlation Matrix for Key Performance Metrics"

# Heatmap correlation
colors <- mako(10)
stops <- seq(0, 1, length.out = length(colors))
custom_colorscale <- lapply(seq_along(colors), function(i) {
  list(stops[i], colors[i])
})

num_vars <- game_data_clean %>%
  select(
    score, game_duration, coins_collected, ufos_shot, 
    bullets_fired, pipes_passed, accuracy
  )

cor_matrix <- cor(num_vars, use = "complete.obs")

plot_ly(
  x = colnames(cor_matrix),
  y = colnames(cor_matrix),
  z = cor_matrix,
  type = "heatmap",
  colorscale = custom_colorscale,
  xgap = 2,
  ygap = 2,
  showscale = TRUE
) %>%
  layout(
    title = list(text = "Correlation Matrix for Key Performance Metrics", x = 0.5),
    xaxis = list(title = "", tickangle = 45, tickfont = list(size = 10)),
    yaxis = list(title = "", autorange = "reversed", tickfont = list(size = 10)),
    plot_bgcolor = "white",
    paper_bgcolor = "white",
    margin = list(l = 80, r = 80, b = 80, t = 80)
  )
```

@fig-2-1-3

### 2.2 Death Reason Deep-Dive

```{r}
#| fig-height: 5
#| fig-cap: "Survival Timeline by Death Reason"
#| label: fig-2-2-1

# Survival Timeline by Death Reason
format_bin <- function(x) {
  x <- gsub("\\(", "", x)
  x <- gsub("\\]", "", x)
  x <- gsub("\\[", "", x)
  x <- gsub("\\)", "", x)
  x <- gsub(",", "-", x)
  x
}

game_data_binned <- game_data_clean %>%
  mutate(duration_bin = cut(game_duration,
                            breaks = seq(0, 140, by = 5),
                            include.lowest = TRUE)) %>%
  filter(!is.na(duration_bin)) %>%
  mutate(duration_label = format_bin(as.character(duration_bin))) %>%
  count(duration_label, death_reason, name = "count")

duration_levels <- format_bin(as.character(levels(cut(seq(0, 100, by = 5),
                                                      breaks = seq(0, 140, by = 5),
                                                      include.lowest = TRUE))))
game_data_binned$duration_label <- factor(game_data_binned$duration_label,
                                          levels = duration_levels)

timeline_plot <- ggplot(
  game_data_binned,
  aes(
    x = duration_label,
    y = count,
    color = death_reason,
    group = death_reason,
    text = paste0(
      "<b>Death Reason:</b> ", death_reason, "<br>",
      "<b>Duration:</b> ", duration_label, " sec<br>",
      "<b>Count:</b> ", count
    )
  )
) +
  geom_line(size = 0.7) +
  geom_point(size = 1.5) +
  labs(
    title = "Survival Timeline by Death Reason",
    x = "Game Duration (seconds)",
    y = "Number of Deaths",
    color = "Death Reason"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, margin = margin(t = 5)),
    axis.text.y = element_text(margin = margin(r = 5))
  )

ggplotly(timeline_plot, tooltip = "text") %>%
  layout(
    title = list(
      text = "<b>Survival Timeline by Death Reason</b>",  
      x = 0.5,          
      xanchor = "center",
      font = list(size = 17)  
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.25,
      yanchor = "top"
    ),
    xaxis = list(
      title_standoff = 20
    ),
    yaxis = list(
      title_standoff = 20
    ),
    margin = list(b = 160)  
  )
```

```{r}
#| fig-height: 5
#| fig-cap: "Score Distribution by Death Reason"
#| label: fig-2-2-2

#  Distribution of score by death_reason
stats <- game_data_clean %>%
  group_by(death_reason) %>%
  summarise(
    count = n(),
    mean  = mean(score),
    min   = min(score),
    q1    = quantile(score, 0.25),
    median= median(score),
    q3    = quantile(score, 0.75),
    max   = max(score)
  )

df <- left_join(game_data_clean, stats, by = "death_reason")

p <- plot_ly()

unique_reasons <- unique(df$death_reason)

for (dr in unique_reasons) {

  dsub <- df %>% filter(death_reason == dr)

  cd <- as.matrix(dsub[, c("count","mean","min","q1","median","q3","max")])

  p <- add_trace(
    p,
    data = dsub,
    x = ~death_reason,
    y = ~score,
    type = "violin",
    name = dr,
    box = list(visible = TRUE),
    meanline = list(visible = TRUE),
    customdata = cd,
    hovertemplate = paste(
      "<b>Death reason:</b> ", dr, "<br>",
      "<b>Score:</b> %{y}<br><br>",
      "<b>Count:</b> %{customdata[0]}<br>",
      "<b>Mean:</b> %{customdata[1]:.2f}<br>",
      "<b>Min:</b> %{customdata[2]}<br>",
      "<b>Q1:</b> %{customdata[3]}<br>",
      "<b>Median:</b> %{customdata[4]}<br>",
      "<b>Q3:</b> %{customdata[5]}<br>",
      "<b>Max:</b> %{customdata[6]}<extra></extra>"
    )
  )
}

 p %>% layout(
  title = "Score Distribution by Death Reason",
  xaxis = list(title = "Death Reason"),
  yaxis = list(title = "Score")
)
```

```{r}
#| label: tbl-2-2-1
#| tbl-cap: "Expected Value of Score per Death Reason"

# Expected Value of Score Lost per Death Type
ev_loss <- game_data_clean %>%
    group_by(death_reason) %>%
    rename(`Death reason` = death_reason) %>%
    summarise(
        `Mean score` = mean(score),
        `Median score` = median(score),
        `Count of deaths` = n(),
        .groups = 'drop'
    ) %>%
    arrange(desc(`Mean score`))

ev_loss %>% kable()
```

### 2.3 Behavioral Feature Engineering

```{r}
# Behavioral features created 
game_data_enhanced <- game_data_clean %>%
    mutate(
        aggressiveness = ifelse(game_duration > 0, bullets_fired / game_duration, 0),
        efficiency = ifelse(game_duration > 0, score / game_duration, 0),
        risk_taking = ifelse(pipes_passed > 0, ufos_shot / pipes_passed, 0),
        coin_rate = ifelse(game_duration > 0, coins_collected / game_duration, 0),
        ufo_rate = ifelse(game_duration > 0, ufos_shot / game_duration, 0),
        pipe_pass_rate = ifelse(game_duration > 0, pipes_passed / game_duration, 0),
        complexity = bullets_fired + ufos_shot + pipes_passed,
        skill_tier = cut(score, 
                        breaks = c(-1, 4, 9, 14, 29, 49, Inf),
                        labels = c("0-4", "5-9", "10-14", "15-29", "30-49", "50+"))
    )
```

```{r}
#| fig-height: 5
#| fig-cap: "Behavioral Features Correlation Matrix"
#| label: fig-2-3-1

# Behavioral features
behavioral_vars <- game_data_enhanced %>%
  select(aggressiveness, efficiency, accuracy, risk_taking,
         coin_rate, ufo_rate, pipe_pass_rate, complexity)

# Correlation matrix
behavioral_cor <- cor(behavioral_vars, use = "complete.obs")

# Heatmap Plotly 
plot_ly(
  x = colnames(behavioral_cor),
  y = colnames(behavioral_cor),
  z = behavioral_cor,
  type = "heatmap",
  colorscale = "RdBu",   
  zmin = -1,
  zmax = 1,
  xgap = 2,
  ygap = 2,
  showscale = TRUE
) %>%
  layout(
    title = list(text = "Behavioral Features Correlation Matrix", x = 0.5),
    xaxis = list(title = "", tickangle = 45, tickfont = list(size = 10)),
    yaxis = list(title = "", autorange = "reversed", tickfont = list(size = 10)),
    plot_bgcolor = "white",
    paper_bgcolor = "white",
    margin = list(l = 80, r = 80, b = 80, t = 80)
  )
```

### 2.4. Session Progression Analysis

```{r}
#| fig-height: 5
#| fig-cap: "Score vs Game Duration with Trendline"
#| label: fig-2-4-1

# (A) Score vs Duration with trendline
score_duration_plot <- ggplot(game_data_enhanced, 
                            aes(x = game_duration, y = score)) +
    geom_point(alpha = 0.6, color = "#1f77b4") +
    geom_smooth(method = "loess", color = "#ff7f0e", se = TRUE) +
    labs(title = "Score vs Game Duration with Trendline",
         x = "Game Duration (seconds)",
         y = "Score") +
    theme_minimal()

ggplotly(score_duration_plot)
```

```{r}
#| fig-height: 5
#| fig-cap: "Bullets Fired vs UFOs Shot"
#| label: fig-2-4-2

# (B) Bullets vs UFO Shot 
efficiency_plot <- ggplot(game_data_enhanced,
                         aes(x = bullets_fired, y = ufos_shot, color = skill_tier)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = "Bullets Fired vs UFOs Shot",
         x = "Bullets Fired", 
         y = "UFOs Shot",
         color = "Skill Tier (by Score)") +
    theme_minimal()

ggplotly(efficiency_plot)
```

### 2.5 Clusterable Structure Check

```{r}
#| fig-height: 5
#| fig-cap: "PCA - Player Behavior Patterns by Death Reason"
#| label: fig-2-5-1

# PCA Analysis
cluster_data <- game_data_enhanced %>%
    select(score, game_duration, coins_collected, bullets_fired,
           ufos_shot, pipes_passed, aggressiveness, efficiency,
           accuracy, risk_taking) %>%
    scale()

pca_result <- prcomp(cluster_data, scale. = TRUE)

pca_df <- as.data.frame(pca_result$x)
pca_df$death_reason <- game_data_enhanced$death_reason
pca_df$score <- game_data_enhanced$score
pca_df$game_duration <- game_data_enhanced$game_duration
pca_df$accuracy <- round(game_data_enhanced$accuracy, 3)
pca_df$coins_collected <- game_data_enhanced$coins_collected
pca_df$ufos_shot <- game_data_enhanced$ufos_shot
pca_df$pipes_passed <- game_data_enhanced$pipes_passed

pca_df$tooltip <- paste(
    "Death reason:", pca_df$death_reason, "<br>",
    "Score:", pca_df$score, "<br>",
    "Duration:", pca_df$game_duration, "s<br>",
    "Accuracy:", pca_df$accuracy, "<br>",
    "Coins:", pca_df$coins_collected, "<br>",
    "UFOs:", pca_df$ufos_shot, "<br>",
    "Pipes:", pca_df$pipes_passed
)

pca_plot <- plot_ly(pca_df, x = ~PC1, y = ~PC2, 
                   color = ~death_reason,
                   type = 'scatter', mode = 'markers',
                   text = ~tooltip,
                   hoverinfo = 'text',
                   marker = list(size = 8, opacity = 0.7)) %>%
    layout(title = "PCA - Player Behavior Patterns by Death Reason",
           xaxis = list(title = paste0("PC1 (", 
                                     round(100 * pca_result$sdev[1]^2/sum(pca_result$sdev^2), 1), "%)")),
           yaxis = list(title = paste0("PC2 (", 
                                     round(100 * pca_result$sdev[2]^2/sum(pca_result$sdev^2), 1), "%)")))

pca_plot
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "PCA Loadings - Meaning of Principal Components"
#| label: fig-2-5-2

# PCA Loadings - Meaning of Principal components
pca_loadings <- as.data.frame(pca_result$rotation[, 1:2])
pca_loadings$feature <- rownames(pca_loadings)

loadings_plot <- ggplot(pca_loadings, aes(x = PC1, y = PC2, label = feature)) +
    geom_point(size = 3, color = "blue") +
    geom_text_repel(size = 4, max.overlaps = 10) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    labs(title = "PCA Loadings - Meaning of Principal Components",
         x = paste0("PC1 (", round(100 * pca_result$sdev[1]^2/sum(pca_result$sdev^2), 1), "%)"),
         y = paste0("PC2 (", round(100 * pca_result$sdev[2]^2/sum(pca_result$sdev^2), 1), "%)")) +
    theme_minimal()

loadings_plot
```

## 3. Bootstrapping Data for Machine Learning

```{r}
# Sort by time and split data
game_sorted <- game_data_enhanced %>% arrange(start_time)
train_base <- head(game_sorted, 250)
test_holdout <- tail(game_sorted, 50)

# Bootstrap training data
set.seed(123)
bootstrap_size <- 10000
train_bootstrapped <- train_base %>% 
    slice_sample(n = bootstrap_size, replace = TRUE) %>% 
    mutate(is_synthetic = TRUE)

cat("Original Train Size:", nrow(train_base), "\n")
cat("Bootstrapped Train Size:", nrow(train_bootstrapped), "\n")
cat("Holdout Test Size:", nrow(test_holdout), "\n")
```

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Score Distribution: Real vs Bootstrapped Data"
#| label: fig-3

# Compare distribution real vs bootstrapped
compare_plot <- ggplot() +
    geom_density(data = train_base, aes(x = score, color = "Real"), size = 1) +
    geom_density(data = train_bootstrapped, aes(x = score, color = "Bootstrapped"), size = 1, alpha = 0.7) +
    labs(title = "Score Distribution: Real vs Bootstrapped Data",
         x = "Score", y = "Density", color = "Data Type") +
    theme_minimal()

compare_plot
```


## 4. Segmentation (Unsupervised Learning)

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Enhanced Player Segmentation (K-Means)"
#| label: fig-4

# Features for clustering
cluster_features_enhanced <- train_bootstrapped %>% 
    select(score, game_duration, coins_collected, bullets_fired,
           ufos_shot, pipes_passed, aggressiveness, efficiency,
           accuracy, risk_taking)

scaled_features_enhanced <- scale(cluster_features_enhanced)

# KMeans clustering with 3 clusters
set.seed(123)
kmeans_enhanced <- kmeans(scaled_features_enhanced, centers = 3, nstart = 25)
train_bootstrapped$cluster_enhanced <- as.factor(kmeans_enhanced$cluster)

# Visualize clusters
fviz_cluster(kmeans_enhanced, data = scaled_features_enhanced,
            geom = "point", ellipse.type = "convex",
            ggtheme = theme_minimal(),
            main = "Enhanced Player Segmentation (K-Means)")
```

```{r}
#| label: tbl-4
#| tbl-cap: "Cluster profiles"

# Cluster profiles
cluster_profiles <- train_bootstrapped %>%
    group_by(cluster_enhanced) %>%
    rename(`Cluster enhanced` = cluster_enhanced) %>%
    summarise(
        Count = n(),
        `Avg Score` = mean(score),
        `Avg Duration` = mean(game_duration),
        `Avg Coins` = mean(coins_collected),
        `Avg UFOs` = mean(ufos_shot),
        `Avg Bullets` = mean(bullets_fired),
        `Avg Aggressiveness` = mean(aggressiveness),
        `Avg Efficiency` = mean(efficiency),
        `Avg Accuracy` = mean(accuracy),
        .groups = 'drop'
    )

cluster_profiles %>% kable()
```

## 5. Predictive Modeling (Supervised Learning)

### 5.1 Score Forecasting (Regression)

We use a Random Forest model to predict the final score based on gameplay metrics.

```{r}
#| label: model-score-rf
#| message: false

library(randomForest)
library(caret)

# Define features
features <- c("coins_collected", "ufos_shot", "bullets_fired", "game_duration", "pipes_passed")

# Train Random Forest on Bootstrapped Data
rf_model_score <- randomForest(
  as.formula(paste("score ~", paste(features, collapse = "+"))),
  data = train_bootstrapped,
  ntree = 100,
  importance = TRUE
)

# Predict on Holdout Test Set
predictions_rf <- predict(rf_model_score, newdata = test_holdout)

# Evaluate Performance (RMSE & R-squared)
rmse_val <- RMSE(predictions_rf, test_holdout$score)
r2_val <- R2(predictions_rf, test_holdout$score)

cat("Random Forest Performance on Holdout Set:\n")
cat("RMSE:", round(rmse_val, 2), "\n")
cat("R-Squared:", round(r2_val, 4), "\n")

# Variable Importance Plot
varImpPlot(rf_model_score, main = "Feature Importance for Score Prediction")
```

### 5.2 Survival Analysis (Logistic Regression)

We predict whether a player will survive past a specific "expert" threshold (e.g., 30 seconds). This is a binary classification problem.

```{r}
#| label: model-survival-logistic

# Define 'Survival' as lasting longer than 30 seconds
threshold <- 30

train_bootstrapped <- train_bootstrapped %>%
  mutate(survived_expert = as.factor(ifelse(game_duration > threshold, 1, 0)))

test_holdout <- test_holdout %>%
  mutate(survived_expert = as.factor(ifelse(game_duration > threshold, 1, 0)))

# Train Logistic Regression
# We remove variables that directly calculate duration/score to prevent data leakage, focusing on behavioral counts
log_model <- glm(survived_expert ~ bullets_fired + ufos_shot + coins_collected, 
                 data = train_bootstrapped, 
                 family = "binomial")

# Predict probabilities on Test Set
probs_survival <- predict(log_model, newdata = test_holdout, type = "response")
preds_survival <- ifelse(probs_survival > 0.5, 1, 0)

# Confusion Matrix
conf_matrix <- confusionMatrix(as.factor(preds_survival), test_holdout$survived_expert)

cat("Logistic Regression Accuracy:", round(conf_matrix$overall['Accuracy'], 4), "\n")
print(conf_matrix$table)
```

## 6. Business Insights & Recommendations

Based on the analysis above, we derive the following actionable insights:

### 6.1. Difficulty Balancing:

Observation: The death_reason analysis highlights the most common obstacles (e.g., pipes vs. enemies). If 'pipe' collisions are disproportionately high early in the game, the initial difficulty curve may be too steep.

Recommendation: Adjust the gap size or spawn rate of the leading cause of death in the first 10 seconds of gameplay to improve retention.

### 6.2. Player Segmentation Strategy:

Observation: K-Means clustering identified distinct groups. (Refer to cluster table: e.g., High-duration/low-coin collectors vs. Aggressive shooters).

Recommendation: Introduce targeted rewards.

- For 'Survivors' (High duration, low action): Introduce time-based achievements.

- For 'Shooters' (High bullets/UFOs): Offer weapon skins or visual upgrades for combat milestones.

### 6.3. Predictive Engagement:

Observation: The Random Forest model shows that specific actions (like coins_collected or ufos_shot) are strong predictors of high scores.

Recommendation: Create a tutorial or "Daily Mission" focusing on these high-value actions to teach new players how to achieve higher scores effectively.

### 6.4. Monetization Opportunities:

Observation: Players who survive past the 30-second threshold (analyzed in the Logistic Regression) show higher engagement.

Recommendation: Trigger "Continue?" ads or special offers only after a player has demonstrated this "expert" survival trait, as they are more invested in the session than a player who dies instantly.
